---
title: "Wilcoxon Wars"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE, 
                      include = TRUE, fig.align = "center")

library(prmisc)
library(moments)

source("wilcoxon_vis.R")
```

## Prologue
To explain how p-values work on my [German science blog](www.einglasrotwein.de), I compared the simulated ratings of two burger restaurants with each other. Both restaurants were rated on a scale from 1 - 10. Then, someone on Twitter asked if it wouldn't have been appropriate to use a nonparametric test instead of the t-test I applied, as the ratings were ordinal data. However, as I had drawn fictional data from a normal distribution, I knew it was interval data. Furthermore, I remembered the paper by [Fagerland (2012)](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/1471-2288-12-78) which showed the Wilcoxon test to suck for skewed data with increasing sample size - while the t-test was unaffected.

Okay, to be precise, the Wilcoxon test did not "suck", but rather answered the different question of Prob(X < Y) = .5 instead of if the groups differ in medians. However, I don't think many people know that (I didn't before I read Fagerland's paper) and are - like me - just told in their statistic lectures to calculate a Wilcoxon test when the assumptions for a t-test are not met. This can lead to ... quite unintended results, as we'll see.

For by blog however, I just wanted to prove that the t-test was the appropriate choice for my simulation (#1). At first. However, I was surprised that it didn't fail, even if I converted the data to be plain ugly ordinal (#2). Then, things got out of hand and I turned to the holy war to bring. That. Wilcoxon. Test. Down.

Make sure to read Fagerland's paper for a far more rigorous and systematic approach with theoretical background. Find the code for my simulations in this [GitHub repo](https://github.com/einGlasRotwein/bits-and-pieces). But for now:

This is Sparta.

### The simulation
A population of 10 million ratings is generated for each restaurant. Then, samples of the following sizes (per restaurant) are drawn: 10, 50, 100, 200, 500, 1000, 2000, 3000, 4000. For each sample size, 10,000 samples of the respective size are drawn. Then, the two restaurants are compared with both, a t-test (Welch test) and a Wilcoxon rank sum (Mann Whitney U) test for independent samples. In seven different scenarios, parameters of the underlying populations are varied to examine the robustness of the two tests. In each of the scenarios, the mean of the two populations is the same. In scenarios 1 - 3, the two populations are exactly identical.

## 01 - default
Normal distribution; integers from 1 - 10. Population 1 and 2 are identical.
```{r, out.width = '50%'}
plot_pop01
```

```{r}
plot_out01
```

## 02 - ordinalskal scale
Same populations as in #01, but converted to an ugly ordinal scale according to the following catagories:

| old rating | label     | new rating |
| :--------- | :-------- | :--------- |
| 1 -  4     | poor      | 1          |
| 5 -  8     | good      | 2          |
| 9 - 10     | excellent | 3          |

```{r out.width = '50%'}
set.seed(kyd)
pop1 <- ordinal_burgers(pop1)

plot_pop02
```

```{r}
plot_out02
```

## 03 - extremely skewed lognormal
```{r}
set.seed(kyd)
pop1 <- rlnorm(pop)
```

Extremely skewed lognormal distribution (skewness = `r round(skewness(pop1), 2)`). Not converted to integers this time. Again, both populations are identical.

```{r}
plot_pop03 +
  facet_zoom(xlim = c(0, 15))
```

```{r}
plot_out03
```

## 04 - different SDs normal distribution
```{r}
set.seed(kyd)
pre04_1 <- rnorm(pop, 5.5, .5)
post04_1 <- burger(pre04_1)

pre04_2 <- rnorm(pop, 5.5, 3)
post04_2 <- burger(pre04_2)
```

Different standard deviations between populations. Right after sampling: population 1 `r print_mean_sd(pre04_1)`; population 2 `r print_mean_sd(pre04_2)`. After conversion to integers from 1 - 10: population 1 `r print_mean_sd(post04_1)`; population 2 `r print_mean_sd(post04_2)`.

```{r}
plot_pop04
```

```{r}
plot_out04
```

## 05 - Different lognormal distributions
```{r}
set.seed(kyd)
pop1 <- rlnorm(pop, sdlog = .72)
pop2 <- rlnorm(pop, sdlog = .3)

diff <- mean(pop1) - mean(pop2)
pop2 <- pop2 + diff
```

Similar to Fagerland: Two lognormal distributions with different SDs. Population 2 was shifted to the same mean as population 1. Note that Fagerland had a more elegant approach where the two distributions differed only in SD while mine also have a different skewness. population 1: `r print_mean_sd(pop1)`, skewness = `r round(skewness(pop1), 2)`; population 2: `r print_mean_sd(pop2)`, skewness = `r round(skewness(pop2), 2)`.

```{r}
plot_pop05_01
```

```{r}
plot_pop05_02
```

```{r}
plot_out05
```

## 06 - Two different gamma distributions
```{r}
set.seed(kyd)
pop1 <- rgamma(pop, shape = .24)
pop2 <- rgamma(pop, shape = .42)

diff <- mean(pop1) - mean(pop2)
pop2 <- pop2 + diff
```

Similar to Fagerland: Two lognormal distributions with different SDs. Population 2 was shifted to the same mean as population 1. Note that Fagerland had a more elegant approach where the two distributions differed only in SD while mine also have a different skewness. population 1: `r print_mean_sd(pop1)`, skewness = `r round(skewness(pop1), 2)`; population 2: `r print_mean_sd(pop2)`, skewness = `r round(skewness(pop2), 2)`.

```{r}
plot_pop06_01
```

```{r}
plot_pop06_02
```

```{r}
plot_out06
```

## 07 - Fagerland inspired escalation
```{r}
set.seed(kyd)
pop1 <- rlnorm(pop, sdlog = .72)
pop2 <- rgamma(pop, shape = .42)

diff <- mean(pop1) - mean(pop2)
pop2 <- pop2 + diff
```

Both distributions used Fagerland's paper: A lognormal distribution (population 1) and a gamma distribution (population 2). Population 2 was shifted to the same mean as population 1. Skewness was comparable for both. population 1: `r print_mean_sd(pop1)`, skew = `r round(skewness(pop1), 2)`; population 2: `r print_mean_sd(pop2)`, skew = `r round(skewness(pop2), 2)`.

```{r}
plot_pop07_01
```

```{r}
plot_pop07_02
```


```{r}
plot_out07
```
