---
title: "Feeling incomplete"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why didn't you tell me, R?
Another day in debugging hell. You know when you ask people: "What's wong?" and they respond with: "Nothing.", even though there clearly **is** something wrong? This is how I like to think about missing error messages when there clearly should be one. The issue I encountered with my code today was the following:

```{r}
(not_cool_r <- data.frame(col1 = 1:3, incomplete_colname = 4:6, incomp = 7:9))
```

Now, we can access the column `incomplete_colname` like this:

```{r}
not_cool_r$incomplete_colname
```

But you know what also works?

This:

```{r}
not_cool_r$incomplete_colnam
```

Or this: 

```{r}
not_cool_r$incomplete_colnam
```

In fact ...

```{r eval = FALSE}
not_cool_r$incomplete_colname
not_cool_r$incomplete_colnam
not_cool_r$incomplete_colna
not_cool_r$incomplete_coln
not_cool_r$incomplete_col
not_cool_r$incomplete_co
not_cool_r$incomplete_c
not_cool_r$incomplete_
not_cool_r$incomplete
not_cool_r$incomplet
not_cool_r$incomple
not_cool_r$incompl # gives column incomplete_colname until here
not_cool_r$incomp # gives column incomp from here
not_cool_r$incom # gives NULL from here because it could be one of two cols
not_cool_r$inco
not_cool_r$inc
```

This is the first call that will give us a different output, because it actually matches the name of column `incomp`.

```{r}
not_cool_r$incomp
```

When leaving out more letters, the result will be `NULL`, because this time, `R` doesn't know whether we meant the column `incomplete_colname` or `incomp`.

```{r}
not_cool_r$incom
```

Why is this a problem? In my case, I called a column that wasn't there. This should have given me an error, but instead, `R` used a different column that happened to have the same beginning like the column name I actually wanted to call.

## How to prevent this
Sure, you can get around this with clever naming. However, with various variable amd column names, you might lose track at some point. At least I didn't think that `sub_id` (subject ID) and `sub_i` (subject intercept) would be a problem.

So, a foolproof way is to access the column via `[[]]`, which will give you `NULL`.

```{r}
not_cool_r[["incomplete_colname"]]
not_cool_r[["incomplete_colnam"]]
```

Or use `[ , ]`, which will give you an error (but don't use `[]`, see [here](https://einglasrotwein.github.io/brackets)).

```{r}
not_cool_r[ , "incomplete_colname"]
try(not_cool_r[ , "incomplete_colnam"])
```

For you `tidyverse` kids out there: The standard tidy syntax doesn't do this ...

```{r message = FALSE}
library(tidyverse)
```

```{r}
not_cool_r %>% 
  mutate(new_col = incomplete_colname * 2)

try(
  not_cool_r %>% 
    mutate(new_col = incomplete_colnam * 2)
)
```

... and tibbles will give you a warning.

```{r}
not_cool_r <- tibble(col1 = 1:3, incomplete_colname = 4:6, incomp = 7:9)

not_cool_r$incomplete_colname
not_cool_r$incomplete_colnam
```

Find the `.Rmd` [here](https://github.com/einGlasRotwein/bits-and-pieces/tree/master/debugging_hell).
